\begin{version}{\Acknowledgments}

The unsatisfactory situation of type-directed propagation in \OCaml has
been on our mind for years; a first concrete step to which this work
can be traced to a specific
example\footnote{\url{https://github.com/ocaml/ocaml/issues/7388}} of
inconvenient order of propagation (between pattern and expression in
a let-binding) presented by Andreas Rossberg as part of general
feedback on the use of \OCaml in the \texttt{wasm} reference
interpreter \citep*{rossberg-wasm}. Gabriel Scherer proposed to delay
the resolution of ambiguous constraints and advised Olivier Martinot
as an intern on this topic in Summer
2020~\citep*{frozen-constraints-2021}. Together they
identified the implementation and specification difficulties; their
implementation would only handle simple cases where no incremental
generalization is necessary (in particular it did not support
backpropagation), and they did not have a precise declarative
specification. Gabriel Scherer kept working on this infrequently until
2024, identified the need for incremental generalization and
instantiation, but its implementations and declarative specifications
remained incomplete.

Alistair O'Brien started working on suspended constraints in Fall
2024, coming from work on constraint-based presentations of GADT
checking---suspended constraints could delay the checking of GADT
patterns until the equalities to be introduced are known to be between
rigid types. He proposed an alternative declarative implementation,
and implemented a working prototype of incremental generalization
instantiation, described in the present paper. Gabriel Scherer and
Alistair O'Brien converged to a semantics using unicity conditions as
presented here.

Didier R\'emy started working on suspended constraints in Winter
2024-2025, coming from questions of principal type inference in
presence of modular implicits, in collaboration with Samuel
Vivien. Didier R\'emy implemented a second prototype of delayed
constraint based on semi-unification, which is fairly different from
the prototype described in the present paper. He worked out the
implementation and formal presentation of polytypes (semi-explicit
first-class polymorphism) using our suspended constraints,
which became one of our main examples.

All co-authors contributed to the writing, with Alistair O'Brien doing
the majority of the writing, and also most of the proofs of the
meta-theory.
\end{version}
